Tree establishment in models of forest dynamics: a key priority for further research

### 2023-04-26 manuscript/manuscriptRW.Rmd
reduce abstract to 300 words

Tree establishment is a key process in forest dynamics, particularly in
the context of forest resilience and climate change. Models are pivotal
for assessing long term (i.e., decadal to centennial) forest dynamics,
and they have been in use for more than 50 years. However, an assessment
of their ability to accurately represent tree establishment is lacking.
We assess how well current models capture the overall abundance, species
composition, and mortality of tree establishment.

Using 15 models and model variants built to capture long-term forest
dynamics at the stand, landscape and global levels, we simulate tree
establishment at 200 sites representing large environmental gradients
across central Europe. The results are evaluated against comprehensive
data from unmanaged forests.

Most of the models overestimate ingrowth levels, which is only
compensated by some models with high simulated mortality rates in the
early stages of individual trees dynamics. Simulated species diversity
of ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth levels
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured well by most models across temperature and soil moisture
gradients, whereas ingrowth of *Fagus sylvatica* is generally
underestimated. *Picea abies* ingrowth is generally overestimated,
particularly at colder sites. Increasing complexity in the tree
establishment modules of the models is not related to higher accuracy of
simulated tree establishment. Furthermore, individual model design is
more important than scale (stand, landscape, global) and approach
(empirical, process-based) for accurately capturing tree establishment.

Despite considerable mismatches between simulation results and data in
some instances, it is remarkable that most of the models capture the
essential features of the highly complex process of tree establishment,
while not having been parameterized with such data. We conclude that in
spite of all the uncertainties, much can be gained by evaluating and
refining model establishment processes. This has the potential to render
long-term projections of forest dynamics under changing environmental
conditions that are much more robust.

@@

Overall, models are needed to (1) synthesize existing empirical data and
explore their relationships, (2) assess future tree establishment e.g.
in the context of global change scenarios, and (3) identify the most
important processes that are shaping ecological patterns. Given the
current strategies that are used in models of forest dynamics to
represent tree establishment, model behavior often is prone to problems,
such as very high levels of tree establishment that necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated, which is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations set-up comprise a multitude of species. Some models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g. under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to unrealistic
successional drift in the model, which must be corrected by factors that
are hard to parameterize [e.g., @lischke2006]. A related issue is the
excessive reduction of species diversity due to positive feedback
effects such that eventually just single-species stands remain. This is
sometimes corrected by the incorporation of a low level of seed influx
of all species at all times [@schumacher2006] or by restricting the
number of seeds per species in the seedbank [@lischke2006]. However,
simulated species composition is usually exceedingly sensitive to
assumptions about seed availability, whereas the parameters of such
functions are poorly constrained by field data. Lastly, there is often a
problem with insufficient observational constraints on parameter values
for models that start from very small tree sizes (e.g., 10 cm of height
or even seeds) and track tree development in a process-oriented manner
by considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

@@

Exercises like the one presented here, where the models are operated in
"blind flight" mode, i.e. without the possibility of being tuned towards
capturing the expected patterns, should be repeated. Such benchmarking
exercises should next focus on aspects such as specific model traits and
the ecological formulations of particular (sub-)processes, to better
understand the implications of the assumptions on which the models are
based. Furthermore, the inclusion of a wide range of models with
different scales, type, populations structures, tree establishment
modules and complexity will ensure a wider benefit to the whole modeling
community.

@@

This study complies with the basic rules of reproducibility and has made
available the protocol of the study, the data and all the software that
has led to the findings presented in this paper. The protocol is
available here @díaz-yáñez2022 . The code and data that support the
findings of this study are openly available at zenodo (*link to be
added*) and github (*link to be added*).

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yáñez, Yannek Käber and Harald Bugmann.
Several people on top of the co-authors list have been involved in the
development of this study in several degrees that we would like to thank
here including Martin Gutsch and Giorgio Vacchiano. Funding for the
workshop was provided by COST Action CA19139 PROCLIAS (PROcess-based
models for CLimate Impact Attribution across Sectors), supported by COST
(European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Research Grant ETH-35 18-1). ODY was funded by the Swiss National
Science Foundation project 200020_188882 / 1 . KB received support from
the European Research Council under the European Union's Horizon 2020
research and innovation program (Grant Agreement 101001905). JB
acknowledges funding from the long-term research development project no.
RVO 67985939 from the Czech Academy of Sciences. MM acknowledges
financial support from I-Maestro (Innovative forest management
strategies for a resilient bioeconomy under climate change and
disturbances, grant no. 22035418, 2019--2022) funded by the ERA-NET
Cofund ForestValue. MMina acknowledges funding from the H2020-MSCA-IF
project REINFORCE (Grant No. 891671). KM received funding from the
Project No. CZ.02.1.01/0.0/0.0/16_019/0000803 financed by OP RDE, grant
no. ITMS2014+313011W580 supported by the Integrated Infrastructure
Operational Programme funded by the ERDF. CH acknowledges financial
support from Waldfonds WASIM2100 project.

@@

```{=html}
<!--The relationship was first described by
[@key] for single citation
[@key1; @key2] multiple citation can be separated by semi-colon
[-@key] in order to suppress author name, and just display the year
[see @key1 p 12; also this ref @key2] is alos a valid syntax
-->
```

>>>>

Tree establishment is a key process in forest dynamics, particularly in
the context of forest resilience and climate change. Models are pivotal
for assessing long-term forest dynamics, and they have been in use for
more than 50 years. However, an assessment of their ability to
accurately represent tree establishment is lacking. We assess how well
current models capture the overall abundance, species composition, and
mortality of tree establishment.

Using 15 models built to capture long-term forest dynamics at the stand,
landscape, and global levels, we simulate tree establishment at 200
sites representing large environmental gradients across Central Europe.
The results are evaluated against comprehensive data from unmanaged
forests.

Most of the models overestimate ingrowth levels, which is only
compensated by some models with high simulated mortality rates in the
early stages of individual trees dynamics. Simulated species diversity
of ingrowth matches the observed ranges. Models simulating higher
species diversity at the stand level do not feature higher ingrowth
diversity. The effect of light availability on ingrowth levels is
captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. Increasing complexity in the
tree establishment modules of the models is not related to higher
accuracy of simulated tree establishment. Furthermore, individual model
design is more important than scale (stand, landscape, global) and
approach (empirical, process-based) for accurately capturing tree
establishment.

Despite considerable mismatches between simulation results and data in
some instances, it is remarkable that most of the models capture the
essential features of the highly complex process of tree establishment,
while not having been parameterized with such data. We conclude that
much can be gained by evaluating and refining model establishment
processes. This has the potential to render long-term projections of
forest dynamics under changing environmental conditions that are much
more robust.

@@

Overall, models are needed to (1) synthesize existing empirical data and
explore their relationships, (2) assess future tree establishment e.g.,
in the context of global change scenarios, and (3) identify the most
important processes that are shaping ecological patterns. Given the
current strategies that are used in models of forest dynamics to
represent tree establishment, model behavior often is prone to problems,
such as very high levels of tree establishment that necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated, which is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations set-up comprise a multitude of species. Some models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g. under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to unrealistic
successional drift in the model, which must be corrected by factors that
are hard to parameterize [e.g., @lischke2006]. A related issue is the
excessive reduction of species diversity due to positive feedback
effects such that eventually just single-species stands remain. This is
sometimes corrected by the incorporation of a low level of seed influx
of all species at all times [@schumacher2006] or by restricting the
number of seeds per species in the seedbank [@lischke2006]. However,
simulated species composition is usually exceedingly sensitive to
assumptions about seed availability, whereas the parameters of such
functions are poorly constrained by field data. Lastly, there is often a
problem with insufficient observational constraints on parameter values
for models that start from very small tree sizes (e.g., 10 cm of height
or even seeds) and track tree development in a process-oriented manner
by considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

@@

Exercises like the one presented here, where the models are operated in
"blind flight" mode, i.e., without the possibility of being tuned
towards capturing the expected patterns, should be repeated. Such
benchmarking exercises should next focus on aspects such as specific
model traits and the ecological formulations of particular
(sub-)processes, to better understand the implications of the
assumptions on which the models are based. Furthermore, the inclusion of
a wide range of models with different scales, type, populations
structures, tree establishment modules and complexity will ensure a
wider benefit to the whole modeling community.

@@

This study complies with the basic rules of reproducibility and has made
available the protocol of the study, the data and all the software that
has led to the findings presented in this paper. The protocol is
available at @díaz-yáñez2022 . The code and data that support the
findings of this study are openly available at zenodo (*link to be
added*) and github (*link to be added*) [for the review process we have
provided these materials [on this
link](https://polybox.ethz.ch/index.php/s/q5MOB1CLcjARs3H) (password:
DownDataCodeRegen1)].

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yáñez, Yannek Käber and Harald Bugmann.
Funding for the workshop was provided by COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and Technology;
[https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich (Research
Grant ETH-35 18-1). ODY was funded by the Swiss National Science
Foundation project 200020_188882 / 1. KB received support from the
European Research Council under the European Union's Horizon 2020
research and innovation program (Grant Agreement 101001905). JB
acknowledges funding from the long-term research development project no.
RVO 67985939 from the Czech Academy of Sciences. MM acknowledges
financial support from I-Maestro (Innovative forest management
strategies for a resilient bioeconomy under climate change and
disturbances, grant no. 22035418, 2019--2022) funded by the ERA-NET
Cofund ForestValue. MMina acknowledges funding from the H2020-MSCA-IF
project REINFORCE (Grant No. 891671). KM received funding from the
Project No. CZ.02.1.01/0.0/0.0/16_019/0000803 financed by OP RDE, grant
no. ITMS2014+313011W580 supported by the Integrated Infrastructure
Operational Programme funded by the ERDF. CH acknowledges financial
support from Waldfonds WASIM2100 project. We would also like to thank
all the people that helped to make this study possible with
contributions at different stages such as Martin Gutsch and Giorgio
Vacchiano.

@@

```{=html}
```
